Figures for 3-D printed oxygen devices

### 2015-04-13 publishing/latex/paper.tex
minor edits from DTE and title change

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{3D-Printed Microfluidic Oxygen Control Devices - Submission to PLOS ONE}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Martin D. Brennan\textsuperscript{1},
Megan L. Rexius-Hall\textsuperscript{1},
David T. Eddington\textsuperscript{1,*},
\\
\bigskip
\bf{1} Dept of Bioengineering, University of Illinois at Chicago, Chicago, Illinois, USA
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Recently 3D printing has emerged as a method for directly printing complete microfluidic chips, although printing materials have been limited to non-oxygen permeable materials.
We demonstrate the addition of gas permeable PDMS (Polydimethylsiloxane) membranes to 3D printed microfluidic devices as a means to enable oxygen control cell culture studies.
The incorporation of a 3D-printed device and gas-permeable membranes was demonstrated on an expanded 24-well version of our previous entirely PDMS device which was an insert for a 6-well plate.
The direct printing allows integrated distribution channels and device geometries not possible with traditional planar lithography.
With this device, four different oxygen conditions were able to be controlled and six wells were maintained under each oxygen condition. 
We domonstrate enhanced transcrtiption of VEGFA (vascular endothelial growth factor A) gene with decreasing oxygen levels in human lung adenocarcinoma cells.
This is the first 3D-printed device that can be functionalized to control oxygen in cell culture.

@@

Printing is currently limited in choice of substrate compatible with the 3D printing process.
Substrate options include many proprietary formulations which have been successfully used in a variety of applications.
New techniques for using a 3D-printed molds to produce devices\cite{Mcdonald2002,Gregory2013,King2014,Comina2014a} are also being developed, including fugitive ink methods\cite{Therriault2003,Miller2012,Bhargava2015}.
To date, there are no widely available methods or materials to facilitate direct printing of gas permeable materials although this area is being developed\cite{Femmer2014a}.
Microfluidic cell culture devices are most commonly cast in PDMS as it is a convenient material for cell studies due to it's biocompatibility, optical properties, and gas permeability, facilitating oxygen control of cell environments\cite{Sia2003,Brennan2014}.

Oxygen control in cells studies is often overlooked by researchers, but important for mimicking conditions experienced by cells in vivo.
Typically cell culture studies are preformed at 21\% oxygen, atmospheric oxygen conditions, although levels that cells experience in vivo are less than 21\% \cite{Brennan2014}.
For example tumors are generally hypoxic as cancer cells rapidly out grow their vasculature creating a poorly perfused, hypoxic inner region\cite{Kumar2014}.
Studying cancer cells under controlled hypoxic conditions is important in understanding the pathophysiology because research has shown hypoxia may enhance aggressive phenotypes, tumor progression, metastasis, and resistance to therapy.
Hypoxia is known to alter the transcription of many genes which are under the activity of the HIF (hypoxia inducible factor) family of transcriptional factors.
To better study the role of oxygen levels in cancer gene expression a gas controlled culture system is required.

@@

Oxygen was measured with PtOEPK (Pt(II) Octaethylporphine ketone) planar sensors that were fixed to the bottom of a 24-well plate.
Senors were prepared as discussed previously \cite{Sinkala2010}. 
Briefly, polystyrene is dissolved in toluene to form a solution to which PtOEPK is mixed in before spinning coating to a thin layer and allowing the toluene to evaporate.
The intensity of the sensor was measured with via florescence microscopy and correlated to the concentration of oxygen with Stern-Volmer analysis (Fig.~\ref{fig2}).

Gas was pulled through the device with negative pressure as a precaution to avoid bulging or detaching membranes, as well as bubble formation in the culture media.
The outlet ports were connected to a water aspirator which pulled gas through the device from the inlet ports.
The inlet tubing was placed in an open cone that was flooded with gas from a tank with a flow that exceed the vacuum flow so that the gas of interest is pulled into the inlet rather then room air due to the open connection.
In this way, the pressure and therefore flow rate is equal across the 4 networks without having to balance flow from 4 pressurized sources, yet different gases of interest can be perfused via vacuum.

@@

\subsection*{Discussion}
This new iteration of the multi-well insert for oxygen control maintains the convenient features of the previous design while improving distribution and interfacing with tubing due to the 3D-printed element.
Oxygen control is preformed in an off-the-shelf culture plate allowing standard protocols to be used for analysis assays or imaging where additional protocols are often required when cells are seeded within microfluidic chips.
The VEGFA gene expression in A549 cells confirms a biological response using the multi-well insert.

@@

\section*{Conclusion}
3D printing microfluidic chips have been limited to non oxygen permeable materials. We demonstrate oxygen control in 3D printed devices the addition of a gas permeable PDMS membrane. 
Oxygen control is demonstrated in a 24 well plate and PCR detection of upregulation of VEGFA mRNA in the A549 shows the device effectively controls oxygen as expected. 
3D printing allows complex designs, integrated tubing connectors, and is comparable in price to standard PDMS fabrication. 
This technique represents a bridge to commercialization where robust devices can be more easily shared and disseminated. 
While injection molding, hot embossing, or other industrial processes are cheaper when making hundreds to thousands of devices, it is not practical to make a injection mold when making tens to hundreds of devices. 
In addition, PDMS fabrication would be too time consuming, expensive, and the failure rate would be unacceptable. 3D printing is a perfect solution to these device fabrication needs.

>>>>

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{3D-printed 24-well plate oxygen control device - Submission to PLOS ONE}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Martin D. Brennan\textsuperscript{1},
Megan L. Rexius-Hall\textsuperscript{1},
David T. Eddington\textsuperscript{1,*},
\\
\bigskip
\bf{1} Dept of Bioengineering, University of Illinois at Chicago, Chicago, Illinois, USA
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Recently 3D printing has emerged as a method for directly printing complete microfluidic devices, although printing materials have been limited to non-oxygen permeable materials.
We demonstrate the addition of gas permeable PDMS (Polydimethylsiloxane) membranes to 3D printed microfluidic devices as a means to enable oxygen control cell culture studies.
The incorporation of a 3D-printed device and gas-permeable membranes was demonstrated on a 24-well oxygen control device for standard multiwell plates.
The direct printing allows integrated distribution channels and device geometries not possible with traditional planar lithography.
With this device, four different oxygen conditions were able to be controlled and six wells were maintained under each oxygen condition. 
We demonstrate enhanced transcription of VEGFA (vascular endothelial growth factor A) gene with decreasing oxygen levels in human lung adenocarcinoma cells.
This is the first 3D-printed device that can be functionalized to control oxygen in cell culture.

@@

Printing is currently limited in choice of substrate compatible with the 3D printing process.
Substrate options include many proprietary formulations which have been successfully used in a variety of applications.
New techniques for using a 3D-printed molds to produce devices\cite{Mcdonald2002,Gregory2013,King2014,Comina2014a} are also being developed, including fugitive ink methods\cite{Therriault2003,Miller2012,Bhargava2015}.
To date, there are no widely available methods or materials to facilitate direct printing of gas permeable materials although this area is actively being explored\cite{Femmer2014a}.
Microfluidic cell culture devices are most commonly cast in PDMS as it is a convenient material for cell studies due to it's biocompatibility, optical properties, and gas permeability, facilitating oxygen control of cell environments\cite{Sia2003,Brennan2014}.

Oxygen control in cells studies is often overlooked by researchers, but important for mimicking conditions experienced by cells in vivo.
Typically cell culture studies are performed at 21\% oxygen, atmospheric oxygen conditions, although levels that cells experience in vivo are less than 21\% \cite{Brennan2014}.
For example tumors are generally hypoxic as cancer cells rapidly outgrow their vasculature creating a poorly perfused, hypoxic inner region\cite{Kumar2014}.
Studying cancer cells under controlled hypoxic conditions is important in understanding the pathophysiology because research has shown hypoxia may enhance aggressive phenotypes, tumor progression, metastasis, and resistance to therapy.
Hypoxia is known to alter the transcription of many genes which are under the activity of the HIF (hypoxia inducible factor) family of transcriptional factors.
To better study the role of oxygen levels in cancer gene expression a gas controlled culture system is required.

@@

Oxygen was measured with PtOEPK (Pt(II) Octaethylporphine ketone) planar sensors that were fixed to the bottom of a 24-well plate.
Sensors were prepared as discussed previously \cite{Sinkala2010}. 
Briefly, polystyrene is dissolved in toluene to form a solution to which PtOEPK is mixed in before spinning coating to a thin layer and allowing the toluene to evaporate.
The intensity of the sensor was measured with via fluorescence microscopy and correlated to the concentration of oxygen with Stern-Volmer analysis (Fig.~\ref{fig2}).

Gas was pulled through the device with negative pressure as a precaution to avoid bulging or detaching membranes, as well as bubble formation in the culture media.
The outlet ports were connected to a water aspirator which pulled gas through the device from the inlet ports.
The inlet tubing was placed in an open cone that was flooded with gas from a tank with a flow that exceed the vacuum flow so that the gas of interest is pulled into the inlet rather than room air due to the open connection.
In this way, the pressure and therefore flow rate is equal across the 4 networks without having to balance flow from 4 pressurized sources, yet different gases of interest can be perfused via vacuum.

@@

\subsection*{Discussion}
This new iteration of the multi-well insert for oxygen control maintains the convenient features of the previous design while improving distribution and interfacing with tubing due to the 3D-printed element.
Oxygen control is performed in an off-the-shelf culture plate allowing standard protocols to be used for analysis assays or imaging where additional protocols are often required when cells are seeded within microfluidic chips.
The VEGFA gene expression in A549 cells confirms a biological response using the multi-well insert.

@@

\section*{Conclusion}
3D printing microfluidic chips have been limited to non oxygen permeable materials. We demonstrate oxygen control in 3D printed devices the addition of a gas permeable PDMS membrane. 
Oxygen control is demonstrated in a 24 well plate and PCR detection of upregulation of VEGFA mRNA in the A549 shows the device effectively controls oxygen as expected. 
3D printing allows complex designs, integrated tubing connectors, and is comparable in price to standard PDMS fabrication. 
This technique represents a bridge to commercialization where robust devices can be more easily shared and disseminated. 
While injection molding, hot embossing, or other industrial processes are cheaper when making hundreds to thousands of devices, it is not practical to make a injection mold when making tens to hundreds of devices. 
In addition, PDMS fabrication would be too time consuming, expensive, and the failure rate would be unacceptable. 3D printing is an excellent solution to these device fabrication needs.
